Investigating the outer reaches of the Sculptor and Ursa Minor dwarf galaxies

### 2025-06-26 thesis/data.tex
thesis w full ch2 draft

To provide an empirical benchmark for simulations in the next chapters,
we aim to understand the observed properties of Sculptor and Ursa Minor.
In this section, we first compile past measurements of each galaxy. We
describe the \citet{jensen+2024} Bayesian methodology to select high
probability member stars from \emph{Gaia}. The derived density profiles
are robust to changes in assumptions. Fornax is well represented by a 2D
exponential profile. Instead, Sculptor and Ursa Minor show an excess of
stars relative to an 2D exponential in the outer regions, better
described my a Plummer profile. Finally, we combine various radial
velocity samples with \citet{jensen+2024}'s members to derive systemic
velocities and dispersions for Sculptor and Ursa Minor.

@@

Since the discovery of dwarf galaxies around the Milky Way,
observational work has attempted to measure and refine the basic
properties of these objects. While the Milky Way's satellites are close
(by extragalactic standards), their low numbers of stars and large areas
on the sky present challenges for observational work. The classical
systems discussed here each extend about 1-2 degrees across the sky
(figs.~\ref{fig:scl_selection}, \ref{fig:umi_selection}, \ref{fig:fornax_selection}).

In particular, \emph{Gaia} has provided a wealth of high quality data,
including the distances to nearby stars and proper motions. This allows
for some of the first measurements on the motions for many local group
dwarf galaxies {[}e.g.; \citet{MV2020}{]}. Accurate distances and
velocities are invaluable for understanding the orbital history of
satellites.

High-quality large spectroscopic samples combined with \emph{Gaia} has
opened new windows into understanding the internal kinematics and
history of dwarf galaxies. \textless Discuss battaglia, pace, etc. and
conclusions about dwarf galaxies.\textgreater{}

In tbls.~\ref{tbl:scl_obs_props}, \ref{tbl:umi_obs_props} we present
adopted observed properties for each galaxy.

\begin{table*}[t]
\centering
\caption[Observed Properties of Sculptor]{Observed properties of Sculptor. References are: 1. Ricardo R. Muñoz et al. (2018), 2. Tran et al. (2022), 3. McConnachie and Venn (2020b), 4 McConnachie and Venn (2020a). }
\label{tbl:scl_obs_props}
\begin{tabular}{lll}
\toprule
parameter & value & Source\\
\midrule
$\alpha$ & $15.0183 \pm 0.0012^\circ$ & 1\\
$\delta$ & $-33.7186 \pm 0.0007^\circ$ & 1\\
distance modulus & $19.60 \pm 0.05$ (RR lyrae) & 2\\
distance & $83.2 \pm 2$ kpc & 2\\
$\mu_{\alpha*}$ & $0.099 \pm 0.002 \pm 0.017$ mas yr$^{-1}$ & 3\\
$\mu_\delta$ & $-0.160 \pm 0.002_{\rm stat} \pm 0.017_{\rm sys}$ mas yr$^{-1}$ & 3\\
radial velocity & $111.3 \pm 0.2\ {\rm km\,s^{-1}}$ & sec. \ref{sec:rv_results}\\
$\sigma_v$ & $9.64\pm0.16\ {\rm km\,s^{-1}}$ & sec. \ref{sec:rv_results}\\
$R_h$ & $9.79 \pm 0.04$ arcmin & 4\\
$R_{h,inner}$ & $9.46 \pm 0.26$ arcmin & XREF\\
ellipticity & $0.36 \pm 0.01$ & 1\\
position angle & $92\pm1^\circ$ & 1\\
$M_V$ & $-10.82\pm0.14$ & 1\\
\bottomrule
\end{tabular}
\end{table*}

\begin{table*}[t]
\centering
\caption[Observed Properties of Ursa Minor]{Observed properties of Ursa Minor. References are: (1) Ricardo R. Muñoz et al. (2018), (2) Garofalo et al. (2025), (3) McConnachie and Venn (2020a), (4) average of Pace et al. (2020) and Spencer et al. (2018). }
\label{tbl:umi_obs_props}
\begin{tabular}{lll}
\toprule
parameter & value & Source\\
\midrule
$\alpha$ & $ 227.2420 \pm 0.0045$˚ & 1\\
$\delta$ & $67.2221 \pm 0.0016$˚ & 1\\
distance modulus & $19.23 \pm 0.11$ (RR lyrae) & 2\\
distance & $70.1 \pm 3.6$ kpc & 2\\
$\mu_\alpha*$ & $-0.124 \pm 0.004 \pm 0.017$ mas yr$^{-1}$ & 3\\
$\mu_\delta$ & $0.078 \pm 0.004_{\rm stat} \pm 0.017_{\rm sys}$ mas yr$^{-1}$ & 3\\
radial velocity & $-245.8 \pm 0.3_{\rm stat} \pm 1_{\rm sys}$ km s$^{-1}$ & sec. \ref{sec:rv_results}\\
$\sigma_v$ & $8.8 \pm 0.2$ & sec. \ref{sec:rv_results}\\
$R_h$ & $11.62 \pm 0.1$ arcmin & 1\\
$R_{h, \rm inner}$ & $11.7\pm0.5$ & \\
ellipticity & $0.55 \pm 0.01$ & 1\\
position angle & $50 \pm 1^\circ$ & 1\\
$M_V$ & $-9.03 \pm 0.05$ & 1\\
\bottomrule
\end{tabular}
\end{table*}

@@

Some of the most fundamental properties of astronomical objects are
their position and velocity. Unfortunately, determining distances to
stars is nontrivial. Additionally, while line-of-sight or radial
velocities (RVs) are easily determined from spectroscopy, the tangental
velocities, perpendicular to RV, are only measurable through proper
motions, typically requiring precise astrometry as well. \emph{Gaia}'s
mission is to produce extremely precise astrometry enabling measurements
of unprecedented accuracy and scale for proper motions, parallaxes, and
magnitudes.

\emph{Gaia} was designed to revolutionize proper motion and parallax
measurements. \emph{Gaia} is a space-based, all-sky survey telescope
with two primary 1.45x0.5m mirrors situated at the Sun-Earth L2 lagrange
point \citep{gaiacollaboration+2016}. \emph{Gaia} was launched in XXX,
completing its mission in 2025 (but with two more data releases
planned). By imagining two patches of sky on the same focal plane,
separated by a fixed angle of 106.5degrees, \emph{Gaia} is able to
measure absolute proper motions by comparing the apparent shifts of
stars in different regions of the sky. In addition to precise
astrometric information, \emph{Gaia} measures the magnitude of stars in
the very wide \emph{G} band (330-1050nm), blue and red colours using the
blue and red photometers (BP and RP, 330-680, 640-1050 respectively),
and takes low resolution BP-RP spectra and radial velocity measurements
of bright stars (magnitudes \textless16?).

\emph{Gaia} has revolutionized many astronomical disciplines, the least
of which is local group and Milky Way science. While proper motions of a
dwarf galaxies has been measured in a case by case basis by the Hubble
Space Telescope, a full systematic determination of proper motions for
most dwarf galaxies was unavailable until \emph{Gaia}
\citep{MV2020a, battaglia+2022}. \emph{Gaia} has furthermore allowed for
the detection of many substructures of the halo, streams,

@@

Here, we briefly describe J+24's membership estimation method. J+24 use
a Bayesian framework incorporating proper motion (PM), colour-magnitude
diagram (CMD), and spatial information to determine the probability that
a given star belongs to the satellite. By accounting for PM in
particular, J+24 produces low contamination samples of candidate member
stars. J+24 extends the algorithm presented in \citet{MV2020a};
\citet{MV2020b} but additionally includes an optional secondary,
extended spatial component to find possible members as far as
\textasciitilde10 half-light radii \(R_h\) from some dwarf galaxies. See
also similar work by \citet{pace+li2019}; \citet{battaglia+2022};
\citet{pace+erkal+li2022}; \citet{qi+2022}.

@@

\begin{itemize}
\tightlist
\item
  Solved astrometry, magnitude, and colour.
\item
  Renormalized unit weight error, \({\rm ruwe} \leq 1.3\), ensuring high
  quality astrometry. \texttt{ruwe} is a measure of the excess
  astrometric noise on fitting a consistent parallax-proper motion
  solution \citep[see][]{lindegrenXXX}.
\item
  3\(\sigma\) consistency of measured parallax with dwarf's distance
  (dwarf parallax is very small; with \citet{lindegren+2021} zero-point
  correction).
\item
  Absolute proper motions, \(\mu_{\alpha*}\), \(\mu_\delta\), less than
  10\(\,{\rm mas\ yr^{-1}}\). (Corresponds to tangental velocities of
  \(\gtrsim 500\) km/s at distances larger than 10 kpc.)
\item
  Corrected colour excess is within expectations:
  \(|C^*| \leq 3\,\sigma_{C^*}(G)\), with \(C^*\) and \(\sigma_{C^*}\)
  from \citet{riello+2021}. Removes stars with unreliable photometry.
\item
  De-reddened \(G\) magnitude is between
  \(22 > G > G_{\rm TRGB} - 5\delta\rm {DM}\). Removes very faint stars
  and stars significantly brighter than the tip of the red giant branch
  (TRGB) magnitude plus the distance modulus uncertainty
  \(\delta {\rm DM}\).
\item
  Colour is between \(-0.5 < {\rm BP - RP} <  2.5\) (dereddened).
\end{itemize}

@@

J+24 define likelihoods \({\cal L}\) representing the probability
density that a star is consistent with either the MW stellar background
(\({\cal L}_{\rm bg}\)) or the satellite galaxy
(\({\cal L}_{\rm sat}\)). In either case, the likelihoods are the
product of a spatial, PM, and CMD term: \begin{equation}
{\cal L} = {\cal L}_{\rm space}\ {\cal L}_{\rm PM}\ {\cal L}_{\rm CMD}.
\end{equation}

Each likelihoods is normalized over their respective 2D parameter space
for both the satellite. Since the likelihoods are normalized,
\(f_{\rm sat}\), representing the fraction of member stars in the field,
controls the relative scaling between the satellite and background
likelihoods. The total likelihood for any star in this model is then
\begin{equation}\protect\phantomsection\label{eq:Ltot}{
{\cal L}_{\rm tot} = f_{\rm sat}{\cal L}_{\rm sat} + (1-f_{\rm sat}){\cal L}_{\rm bg}
}\end{equation} The probability that any star belongs to the satellite
is then given by \begin{equation}
P_{\rm sat} = \frac{f_{\rm sat}{\cal L}_{\rm sat}}{f_{\rm sat}{\cal L}_{\rm sat} + (1-f_{\rm sat}){\cal L}_{\rm bg}}.
\end{equation}

For the satellite's spatial likelihood, J+24 consider both one-component
and a two-component cases. The one component model is constructed as a
single exponential profile (\(\Sigma \propto e^{R_{\rm ell} / R_s}\)),
with \(R_s\) fixed to the value in the table in \citet{MV2020a}.
Additionally, structural uncertainties (for position angle, ellipticity,
and scale radius) are sampled over to construct the final likelihood
map. The two-component model instead adds a second exponential,
\(\Sigma_\star \propto e^{-R/R_s} + B\,e^{-R/R_{\rm outer}}\). The inner
scale radius is fixed, and the outer scale radius and magnitude of the
second component \(R_{\rm outer}\), \(B\) are solved for. Structural
properties are not accounted for.

@@

The satellite's CMD likelihood is based on a Padova isochrone
\citep{girardi+2002}. The isochrone has a matching metallicity and 12
Gyr age. The (gaussian) colour width is assumed to be 0.1 mag plus the
Gaia colour uncertainty at each magnitude. The HB is modelled as a
constant magnitude extending blue of the CMD (mean magnitude of -2.2, 12
Gyr HB stars and a 0.1 mag width plus the mean colour error). A
likelihood map is constructed by sampling the distance modulus in
addition to the CMD width, taking the maximum of RGB and HB likelihoods.

@@

J+24 derive \(\mu_{\alpha*}\), \(\mu_\delta\), \(f_{\rm sat}\) (and
\(B\), \(R_{\rm outer}\) for two-component) through an MCMC simulation
with likelihood from Eq.~\ref{eq:Ltot}. Priors are weakly informative or
uniform. The proper motion single component prior is same as
\citet{MV2020a}: a normal distribution with mean 0 and standard
deviation \(100\,\kms\). If 2-component spatial, instead is a uniform
distribution spanning 5\(\sigma\) of single component case w/ systematic
uncertainties. \(f_{\rm sat}\) (and \(B\)) has a uniform prior 0--1.
\(R_{\rm outer}\) has a uniform prior only restricting
\(R_{\rm outer} > R_s\). The mode of each parameter from the MCMC are
then reported and used to calculate the final \(P_{\rm sat}\) values.

We adopt a probability cut of \(P_{\rm sat} = 0.2\) as our fiducial
sample. Most stars are assigned probabilities close to either 0 or 1, so
the choice of probability threshold is only marginally significant.
Additionally, even for a probability cut of 0.2, the purity of the
resulting sample with RV measurements is very high (\textasciitilde90\%,
J+24). Note there is likely a systematic bias in using stars with RV
measurements to measure purity. Fainter stars are less likely to have
been targeted and have poorer astrometry and colour information. We find
no difference in the resulting density distributions when restricting
stars to be brighter than a specific magnitude.

@@

In
Figs.~\ref{fig:sculptor_selection}, \ref{fig:umi_selection}, \ref{fig:fornax_selection},
we illustrate the resulting samples from the algorithm in the tangent
plane (\(\xi\), \(\eta\), \emph{does this need defined?}), CMD, and
proper motion space. For all galaxies, each criteria plays a
commensurate role in sifting out nonmembers. The CMD is well defined and
probable members only extend a few times the colour uncertainty from the
CMD. In proper motion space, selected stars are within \(\masyr\) from
the systematic proper motion. The stars furthest away in proper motion
space typically have large uncertainties \(\sim 1\masyr\), so most
members are reasonably consistent. In any case, the algorithm does
enable removing many stars which do not have consistent proper motions
with the dwarf. Finally, the spatial likelihood reduces the
probabilities of stars distant from the dwarfs centres. Member stars are
rare outside of \(\sim5R_h\), resulting from both the likelihood
specification and the lower density of background stars consistent with
the CMD and PM of the satellite.

One limitation of the J+24 method is that it assumes a spatial form of
the dwarf galaxy, possibly limiting the detection of very distant stars
or features. Figs.~\ref{fig:sculptor_selection}, \ref{fig:umi_selection}
also illustrates the distribution of stars selected without a spatial
criterion. We define the \emph{CMD+PM} selection as
\begin{equation}\protect\phantomsection\label{eq:sel_cmd_pm}{
{\cal L}_{\rm CMD,\ sat}\ {\cal L}_{\rm PM,\ sat} > {\cal L}_{\rm CMD,\ bg}\ {\cal L}_{\rm PM,\ bg}
}\end{equation} with the likelihoods from J+24 as described above. These
stars are distributed similar to the fiducial (probable members) sample
but with a fainter uniform distribution across the entire field. This
illustrates the approximate background of stars which may be confused as
members. Additionally, since there is no clear spatial structure in the
\emph{CMD+PM} (or \emph{all}) sample outside several \(R_h\), it is
unlikely that there are additional faint, tidal features detectable with
\emph{Gaia}. Not shown here, we also try a variety of simpler, absolute
cuts and thresholds, finding no evidence of other structure in both
\emph{Gaia} and DELVE/UNIONS data.

>>>>

To provide an empirical benchmark for simulations in the next chapters,
we aim to understand the observed properties and density profiles of
Sculptor and Ursa Minor. In this section, we first compile past
measurements of each galaxy. We then describe the \citet{jensen+2024}
Bayesian methodology to select high probability member stars from
\emph{Gaia}. We show the derived density profiles are robust to changes
in assumptions. While Fornax is well represented by a 2D exponential
profile, Sculptor and Ursa Minor show an excess of stars relative to an
2D exponential in the outer regions, better described by a Plummer
profile. Future chapters assess possible explanations for these excess.

@@

Since the discovery of dwarf galaxies around the Milky Way,
observational work has attempted to measure and refine their basic
properties. While the Milky Way's satellites are close (by extragalactic
standards), their low numbers of stars and large apparent sizes present
challenges for observational work. In particular, the properties of the
faintest dwarf galaxy systems and especially \emph{ambiguous} systems
remains a pressing challenge as a unique probe into the nature of the
early universe and dark matter.

Among local dwarf galaxies, classical systems are perhaps the systems
best studied and with the best derived properties.
Tables~\ref{tbl:scl_obs_props}, \ref{tbl:umi_obs_props} present our
adopted observed properties for Sculptor and Ursa Minor. While extending
1-2 degrees across the sky (e.g.
Figs.~\ref{fig:scl_selection}, \ref{fig:umi_selection}, \ref{fig:fornax_selection}),
classical dwarfs larger number of bright (giant) stars allow for
thousands of stars to be observed with deep photometry and spectroscopy
(e.g. \citet{tolstoy+2023}, \citet{pace+2020}). As a result, the
determination of fundamental properties such as the position, size,
orientation, distance (from 100s of RRL stars, e.g. \citet{tran+2022};
\citet{garofalo+2025}), proper motions (from Gaia, described in next
section), line-of-sight (LOS) velocity, and velocity dispersion
\(\sigma_v\) are relatively well constrained today and show excellent
convergence among different works. However, ongoing research still is
refining our understanding of the detailed structure of Milky Way
satellites, hinting that these objects may be more complex than at first
glance.

A number of works have previously speculated that dwarf galaxies such as
Sculptor and Ursa Minor may have signatures of tidal perturbations,
however not without contention. While \citet{hodge1961} and
\citet{demers+krautter+kunkel1980} were some of the earliest work to
derive the density profile of Sculptor, \citet{innanen+papp1979} was
perhaps the first to speculate that Sculptor may harbour a substantial
population of ``extratidal stars'' (stars beyond the tidal radius),
finding candidate members out to 180' in an elongated distribution from
\citet{vanagt1978}`s catalogue of variable stars.\footnote{Interestingly,
  \citet{innanen+papp1979} also speculate about Uranus's satellite
  distribution, covering gravitational dynamics at two very different
  scales.} Additionally, \citet{eskridge1988} showed a possible excess
of stars in Sculptor relative to a \citet{king1962} and Exponential
density profile beginning around 50', but suggest that this excess may
not be unusual. Later work by \citet{IH1995}, \citet{walcher+2003}, and
\citet{westfall+2006} also showed evidence of an extended component of
Sculptor's density profile, interpreting these stars as either evidence
of tidal debris or a dwarf galaxy halo. On the other hand,
\citet{coleman+dacosta+bland-hawthorn2005} show that Sculptor is well
described by a two component density profile, they additionally mention
that it is unlikely many of Sculptor's stars (less than 2\%) are
extratidal. While \citet{munoz+2018} is the most recent and deepest
photometric study of Sculptor, they only cover an area of out to radii
of about 30', so they are unable to study the possible extended
components of Sculptor detected in past works. In addition to the
presence of ``extratidal stars'', many studies note that Sculptor's
stars appear to become more elliptical with radius, consistent with
tidal effects (\citet{IH1995}; \citet{westfall+2006}).

Ursa Minor was often noted not for a density excess but for unusual
features. Starting with the first density profile from
\citet{hodge1964}, Ursa Minor is known to be highly elliptical. Studies
by \citet{olszewski+aaronson1985}, \citet{demers+1995}, \citet{IH1995},
\citet{kleyna+1998}, and \citet{bellazzini+2002} note that Ursa Minor
appears to contain substructure along its major axis, but without strong
interpretation for the causes. As one interpretation \citet{kleyna+2003}
suggest that this feature is a long-lived star cluster residing in a
cored dark matter halo to be dynamically stable. One of the first direct
suggestions of tidal features came from \citet{martinez-delgado+2001},
who find that stars extend far beyond the nominal tidal radius for Ursa
Minor in the direction of the galaxies elongation, interpreting this as
evidence for tidal effects. \citet{palma+2003} corroborate many of these
earlier findings, showing further evidence for Ursa Minor's peculiar
morphology including S-shaped contours, a possible extended extratidal
population of stars, and a failure for a King density profile to
adequately capture the actual stellar distribution. Each of these
characteristics appear to indicate strong tidal disruption. Using
velocity observations, \citet{pace+2014} additionally show that there
are multiple components in spatial-velocity information. However,
\citet{munoz+2018}'s more modern photometry show a more regular
distribution of stars. Irregardless, Ursa Minor has had enough work
suggesting peculiarities that a deeper investigation into the
possibility of tidal effects is worthwhile.

However, everything discussed so far is before \emph{Gaia}. Thus, the
knowledge of the orbits of these systems is largely unknown and most
surveys could only filter out stars using photometry. As a result, the
tidal radius could only be guessed based on either the density profile
or current distance of the dwarf galaxy. With \emph{Gaia} (discussed
more below), recent work has used Bayesian frameworks to derive
systematic proper motions of many dwarf galaxies (e.g. \citet{MV2020a})
and filter away foreground stars using proper motions and parallax to
detect distant members, and study the 6D internal kinematics of these
galaxies (e.g. \citet{tolstoy+2023}). Most relevant here,
\citet{jensen+2024} present a bayesian algorithm to determine likely
members in \emph{Gaia} (described below), and detect extended secondary
components for 9 dwarf galaxies, including Sculptor and Ursa Minor.
\citet{sestito+2023a} and \citet{sestito+2023b} followed up a few of the
most distant stars detected in \citet{jensen+2024} confirming that
members exist in each galaxy as far ass 9-12 \(R_h\) from the centres.
In this chapter, we discuss the origin and reliability of these
detections, confirming that the density profiles are indeed robust.
These density profiles then provide the scientific motivation to explore
the tidal interpretation in the next chapters.

\emph{Maybe define \(R_{\rm ell}\)} \emph{and} \(\xi, \eta\)
\emph{here}? \emph{We will discuss multiple populations in future
chapters}

\begin{table*}[t]
\centering
\caption[Observed Properties of Sculptor]{Observed properties of Sculptor. References are: 1. Ricardo R. Muñoz et al. (2018) Sérsic fits, 2. Tran et al. (2022), 3. McConnachie and Venn (2020b), 4. Arroyo-Polonio et al. (2024), 5. McConnachie and Venn (2020a). }
\label{tbl:scl_obs_props}
\begin{tabular}{lll}
\toprule
parameter & value & Source\\
\midrule
$\alpha$ & $15.0183 \pm 0.0012^\circ$ & 1\\
$\delta$ & $-33.7186 \pm 0.0007^\circ$ & 1\\
distance modulus & $19.60 \pm 0.05$ (RR lyrae) & 2\\
distance & $83.2 \pm 2$ kpc & 2\\
$\mu_{\alpha*}$ & $0.099 \pm 0.002 \pm 0.017$ mas yr$^{-1}$ & 3\\
$\mu_\delta$ & $-0.160 \pm 0.002_{\rm stat} \pm 0.017_{\rm sys}$ mas yr$^{-1}$ & 3\\
LOS velocity & $111.2 \pm 0.3\ {\rm km\,s^{-1}}$ & 4\\
$\sigma_v$ & $9.7\pm0.2\ {\rm km\,s^{-1}}$ & 4\\
$R_h$ & $9.79 \pm 0.04$ arcmin & 1\\
ellipticity & $0.37 \pm 0.01$ & 1\\
position angle & $94\pm1^\circ$ & 1\\
$M_V$ & $-10.82\pm0.14$ & 1\\
\bottomrule
\end{tabular}
\end{table*}

\begin{table*}[t]
\centering
\caption[Observed Properties of Ursa Minor]{Observed properties of Ursa Minor. References are: (1) Ricardo R. Muñoz et al. (2018) Sérsic fits, (2) Garofalo et al. (2025), (3) McConnachie and Venn (2020a), (4) Pace et al. (2020), average of MMT and Keck results. }
\label{tbl:umi_obs_props}
\begin{tabular}{lll}
\toprule
parameter & value & Source\\
\midrule
$\alpha$ & $ 227.2420 \pm 0.0045$˚ & 1\\
$\delta$ & $67.2221 \pm 0.0016$˚ & 1\\
distance modulus & $19.23 \pm 0.11$ (RR lyrae) & 2\\
distance & $70.1 \pm 3.6$ kpc & 2\\
$\mu_\alpha*$ & $-0.124 \pm 0.004 \pm 0.017$ mas yr$^{-1}$ & 3\\
$\mu_\delta$ & $0.078 \pm 0.004_{\rm stat} \pm 0.017_{\rm sys}$ mas yr$^{-1}$ & 3\\
LOS velocity & $-245.9 \pm 0.3_{\rm stat} \pm 1_{\rm sys}$ km s$^{-1}$ & 4\\
$\sigma_v$ & $8.6 \pm 0.3$ & 4\\
$R_h$ & $11.62 \pm 0.1$ arcmin & 1\\
ellipticity & $0.55 \pm 0.01$ & 1\\
position angle & $50 \pm 1^\circ$ & 1\\
$M_V$ & $-9.03 \pm 0.05$ & 1\\
\bottomrule
\end{tabular}
\end{table*}

@@

Some of the most fundamental properties of astronomical objects are
their positions and velocities. Unfortunately, determining distances to
stars is nontrivial. Additionally, while line-of-sight (LOS) velocities
are easily determined from spectroscopy, the tangental velocities,
perpendicular to the LOS, are best measured through proper motions,
typically requiring accurate determinations of small changes in a star's
position. \emph{Gaia}'s mission is to produce extremely precise
astrometry enabling measurements of unprecedented accuracy and scale for
proper motions, parallaxes, and magnitudes.

\emph{Gaia} was designed to revolutionize proper motion and parallax
measurements. \emph{Gaia} is a space-based, all-sky survey telescope
with two primary 1.45x0.5m mirrors situated at the Sun-Earth L2 Lagrange
point \citep{gaiacollaboration+2016}. \emph{Gaia} was launched in 2013,
completing its mission in 2025 (but with two more data releases
planned). By imagining two patches of sky on the same focal plane,
separated by a fixed angle of 106.5 degrees, \emph{Gaia} is able to
measure absolute proper motions. Stars in different regions of the sky
are affected by parallax differently, so by observing the relative
positions of stars separated by large angles during multiple epochs over
a year, an absolute all-sky reference frame can be derived
\citep{gaiacollaboration+2016}. In addition to precise astrometric
information, \emph{Gaia} measures the magnitude of stars in the very
wide \emph{G} band (330-1050nm), blue and red colours using the blue and
red photometers (BP and RP, 330-680, 640-1050 respectively), and takes
low resolution BP-RP spectra and radial velocity measurements of bright
stars (\emph{Gaia} radial velocity magnitudes \textless16)
\citep{gaiacollaboration+2016}. For our purposes, most useful are
\emph{Gaia}'s measurements of \(G\) magnitude,
\(G_{\rm BP} - G_{\rm RP}\) colour, the position \(\alpha, \delta\), and
the proper motions in RA and declination \(\mu_{\alpha*}\) and
\(\mu_\delta\) (where \(\mu_{\alpha*} = \mu_\alpha \cos \delta\)
corrects for projection effects).

\emph{Gaia} has revolutionized many astronomical disciplines, the least
of which is local group and Milky Way science. While proper motions of a
dwarf galaxies has been measured in a case by case basis by the Hubble
Space Telescope, a full systematic determination of proper motions for
most dwarf galaxies was unavailable until \emph{Gaia}
\citep{MV2020a, battaglia+2022}. \emph{Gaia} has furthermore allowed for
the detection and improved measurements of halo substructures
(CITATIONS), streams (CITAITONS), and Milky Way structure and
kinematics, among many other applications.

@@

\subsection{Quality cuts and membership
model}\label{quality-cuts-and-membership-model}

Here, we briefly describe \citet{jensen+2024}'s (hereafter J+24)
membership estimation method. J+24 use a Bayesian framework
incorporating proper motion (PM), colour-magnitude diagram (CMD), and
spatial information to determine the probability that a given star
belongs to the satellite. By accounting for PM in particular, J+24
produces low contamination samples of candidate member stars out to
large distances from a dwarf galaxy. J+24 extends the algorithm
presented in \citet{MV2020a}; \citet{MV2020b} but additionally includes
an optional secondary, extended spatial component to find possible
members as far as \textasciitilde10 half-light radii \(R_h\) from
several dwarf galaxies. See also similar work by \citet{pace+li2019};
\citet{battaglia+2022}; \citet{pace+erkal+li2022}; \citet{qi+2022}.

@@

\begin{itemize}
\tightlist
\item
  Solved astrometry, magnitude, and colour.
\item
  Renormalized unit weight error, \({\rm ruwe} \leq 1.3\), ensuring high
  quality astrometry. \texttt{ruwe} is a measure of the excess
  astrometric noise on fitting a consistent parallax-proper motion
  solution \citep[see][]{lindegren+2021}.\\
\item
  3\(\sigma\) consistency of measured parallax with dwarf's distance
  (dwarf parallax is very small; with \citet{lindegren+2021} zero-point
  correction).
\item
  Absolute proper motions, \(\mu_{\alpha*}\), \(\mu_\delta\), less than
  10\(\,{\rm mas\ yr^{-1}}\). (Corresponds to tangental velocities of
  \(\gtrsim 500\) km/s at distances larger than 10 kpc.)
\item
  Corrected colour excess is within 3\(\sigma\) of the expected
  distribution from \citet{riello+2021}. Removes stars with unreliable
  photometry.
\item
  De-reddened \(G\) magnitude is between
  \(22 > G > G_{\rm TRGB} - 5\sigma_{\rm DM}\). Removes very faint stars
  and stars significantly brighter than the tip of the red giant branch
  (TRGB) magnitude plus the distance modulus uncertainty
  \(\sigma_{\rm DM}\).
\item
  Colour is between \(-0.5 < {\rm BP - RP} <  2.5\) (dereddened).
  Removes stars substantially outside the expected CMD.
\end{itemize}

@@

J+24 define likelihoods \({\cal L}\) representing the probability
density that a star is consistent with either the MW stellar background
(\({\cal L}_{\rm bg}\)) or the satellite galaxy
(\({\cal L}_{\rm sat}\)). In either case, the likelihoods are the
product of a spatial, PM, and CMD term: \begin{equation}{
{\cal L} = {\cal L}_{\rm space}\ {\cal L}_{\rm PM}\ {\cal L}_{\rm CMD}.
}\end{equation}

Each likelihood is normalized over their respective 2D parameter space
for both the satellite. To control the relative frequency of member and
background stars, \(f_{\rm sat}\) representing the fraction of member
stars in the field. The total likelihood for any star in this model is
the sum of the satellite and background likelihoods, weighted by their
relative frequencies,
\begin{equation}\protect\phantomsection\label{eq:Ltot}{
{\cal L}_{\rm tot} = f_{\rm sat}{\cal L}_{\rm sat} + (1-f_{\rm sat}){\cal L}_{\rm bg}.
}\end{equation} The probability that any star belongs to the satellite
is then given by \begin{equation}{
P_{\rm sat} = 
\frac{f_{\rm sat}\,{\cal L}_{\rm sat}}{{\cal L}_{\rm tot}}
= \frac{f_{\rm sat}{\cal L}_{\rm sat}}{f_{\rm sat}{\cal L}_{\rm sat} + (1-f_{\rm sat}){\cal L}_{\rm bg}}.
}\end{equation}

For the satellite's spatial likelihood, J+24 consider both one-component
and a two-component density models. The one component model is
constructed as a single exponential profile ( surface density
\(\Sigma \propto e^{R_{\rm ell} / R_s}\)), with scale radius \(R_s\)
fixed to the value in table 1 of \citet{MV2020a} from \citet{munoz+2018}
(for a Sérsic fit). Additionally, structural uncertainties (for position
angle, ellipticity, and scale radius) are sampled over to construct the
final likelihood map. The two-component model instead adds a second
exponential,
\(\Sigma_\star \propto e^{-R/R_s} + B\,e^{-R/R_{\rm outer}}\). The inner
scale radius is fixed, and the outer scale radius and magnitude of the
second component \(R_{\rm outer}\), \(B\) are free parameters.
Structural property uncertainties are not included in the two-component
model.

@@

The satellite's CMD likelihood is based on a Padova isochrone
\citep{girardi+2002}. The isochrone has a matching metallicity and 12
Gyr age (except 2 Gyr is used for Fornax). The (gaussian) colour width
is assumed to be 0.1 mag plus the Gaia colour uncertainty at each
magnitude. The horizontal branch is modelled as a constant magnitude
extending blue of the CMD (mean magnitude of -2.2, 12 Gyr HB stars and a
0.1 mag width plus the mean colour error). A likelihood map is
constructed by sampling the distance modulus in addition to the CMD
width, taking the maximum of RGB and HB likelihoods.

@@

J+24 derive \(\mu_{\alpha*}\), \(\mu_\delta\), \(f_{\rm sat}\) (and
\(B\), \(R_{\rm outer}\) for two-component) through an MCMC simulation
with likelihood from Eq.~\ref{eq:Ltot}. Priors are only weakly
informative. The proper motion single component prior is same as
\citet{MV2020a}: a normal distribution with mean 0 and standard
deviation \(100\ \kms\). If 2-component spatial, instead is a uniform
distribution spanning 5\(\sigma\) of single component case w/ systematic
uncertainties. \(f_{\rm sat}\) (and \(B\)) has a uniform prior 0--1.
\(R_{\rm outer}\) has a uniform prior only restricting
\(R_{\rm outer} > R_s\). The mode of each parameter from the MCMC are
then reported and used to calculate the final \(P_{\rm sat}\) values.

@@

In
Figs.~\ref{fig:scl_selection}, \ref{fig:umi_selection}, \ref{fig:fornax_selection},
we illustrate the resulting samples from the algorithm in the tangent
plane, CMD, and proper motion space. The tangent plane coordinates
\(\xi\) and \(\eta\) are the distances in RA and declination as measured
on a plane tangent to the dwarf galaxies centre.

For our fiducial sample, we adopt a probability cut of
\(P_{\rm sat} = 0.2\). Most stars are assigned probabilities close to
either 0 or 1, so the choice of probability threshold is only marginally
significant. Additionally, even for a probability cut of 0.2, the purity
of the resulting sample with RV measurements is very high
(\textasciitilde90\%, J+24). Note there is likely a systematic bias in
using stars with RV measurements to measure purity. Fainter stars are
less likely to have been targeted and have poorer astrometry and colour
information. We find no difference in the resulting density
distributions when restricting stars to be brighter than a specific
magnitude.

For all galaxies, each criteria is commensurate in sifting out
nonmembers. The CMD is well defined, and probable members only extend a
few times the colour uncertainty from the CMD. In proper motion space,
selected stars are within \(\masyr\) from the systematic proper motion.
The stars furthest away in proper motion space typically have large
uncertainties \(\sim 1\masyr\), so most members are reasonably
consistent. In any case, the algorithm removes stars with inconsistent
PMs. Finally, the spatial likelihood reduces the probabilities of stars
distant from the dwarfs centres. Member stars are rare outside of
\(\sim5R_h\), resulting from both the likelihood specification and the
lower density of background stars consistent with the CMD and PM of the
satellite.

One limitation of the J+24 method is that it assumes a spatial form of
the dwarf galaxy, possibly limiting the detection of very distant stars
or features. Figs.~\ref{fig:scl_selection}, \ref{fig:umi_selection} also
illustrates the distribution of stars selected without a spatial
criterion. We define the \emph{CMD+PM} selection as
\begin{equation}\protect\phantomsection\label{eq:sel_cmd_pm}{
{\cal L}_{\rm CMD,\ sat}\ {\cal L}_{\rm PM,\ sat} > {\cal L}_{\rm CMD,\ bg}\ {\cal L}_{\rm PM,\ bg}
}\end{equation} with the likelihoods from J+24 as described above. These
stars are distributed similar to the fiducial (probable members) sample
but with a background uniform distribution across the entire field. This
illustrates the approximate background of stars which may be confused as
members. Additionally, since there is no clear spatial structure in the
\emph{CMD+PM} (or \emph{all}) sample outside several \(R_h\), it is
unlikely that there are additional faint, tidal features detectable with
\emph{Gaia}. Not shown here, we also try a variety of simpler, absolute
cuts and thresholds, finding no evidence of other structure in both
\emph{Gaia} and DELVE/UNIONS data.
